---
pagetitle: 'Learning Coding in R'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Basic Plotting {.tabset}

In *R*, you can choose different ways to plot your data and results. By default, *R* comes with built-in graphical functions. Other functions, or entirely different paradigms, can be enabled using different packages. Here we give a brief overview of the standard plotting functions that are useful for quick data exploration. These are also referred to as R “base graphics” as no extra package is required.

### Scatter Plots

The function ```plot(x, y, ...)``` is the most basic way to plot points defined by their x- and y-coordinates. Here, *x* and *y* are vectors of equal length containing the x- and y-coordinates of the points to be plotted. For example,

```{r eval=FALSE}
# generate data
x <- 1:30 # creates a vector of 30 values that we use as x-coordinates
y <- rnorm(30, mean = x) # creates a vector of 30 values that we use as y-coordinates; each values is drawn from normal distribution with with an increasing mean value
plot(y) # if no x-vector is given the default x = (1, 2, 3, ...) is used
plot(x, y) # this plot produces an identical result but strictly speaking x is not necessary here
# there are many different option one can add to the plot command
plot(y, type = "l") # using lines instead of points
plot(y, type = "b") # using both lines and points
plot(y, type = "b", pch = 4) # change the type of point
plot(y, type = "b", pch = 4, col = "blue") # change color
abline(a = 0, b = 1) # add a line to the same plot by specifying its intercept a and slope b
abline(h = 20) # adding a horizontal line by setting a y-value h
abline(v = 10) # adding a vertical line by setting an x-value v
plot(y, col = "orange", xlab = "my x-label", ylab = "a y-label") # add axes labels
plot(y, xlim = c(1,35), ylim = c(1,40)) # set plot ranges
```

In the above examples, we plotted the entries of a vector *v* without specifying values for the x-axes of the plot. Indeed, if no x-vector is specified, then the default *x = (1, 2, 3, ...)* is used. This does not work anymore if y-values do not correspond to x-values that are consecutive natural numbers. Here an example:

```{r eval=FALSE}
x <- seq(0, 5, 0.5) 
x
y <- x^2
plot(x, y) # check the x-axes; this is the desired plot
plot(y) # without specifying x, the y-values are at wrong positions
```

*R* base graphics are drawn interactively: you can start with a plot, and then overlay other elements on the existing plot.

```{r eval=FALSE}
x <- 1:30
y <- rnorm(30, mean = x)

plot(x, y, type = "l", ylim = c(1,60), col = "orange", ylab = "f(x)") 
lines(x, 2*y, type = "l", col = "blue") # adds a line to the previous plot
legend("topleft", legend = c("f(x)=y","f(x)=2y"), lty = c(1,1), col = c("orange", "blue")) # adds a plot legend
```


### Function Plots

We can plot the graphs of functions using the command ```curve(expr, from, to, ...)```, where *expr* is a mathematical expression containing an argument *x* or function name, and *from* and *to* defines the range of *x*-values for which the function is plotted. This is a useful tool to visually explore the shape of functions.

```{r eval=FALSE}
curve(exp, from = -3, to = 3) # plot of the exponential function in the interval (-3, 3)
abline(v = 0) # adding a vertical line at x = 0
logistic <- function(n, r, k){# definition of the user-defined function giving the population growth rate of a logistically growing population as a function the current population size n, the intrinsic growth rate r and the carrying capacity k
  n*r*(1-n/k)
}

curve(logistic(x , r = 1, k = 2), from = 0, to =2.5) # plot of the function logistic as a function of the population size n = x
# note that the function curve requires that you name the argument that you use for plotting x
abline(h = 0) # adding a horizontal line at y = 0
```

### Plotting Matrix Columns

The plotting command ```matplot(x, y, ...)``` is an easy way to superimpose several data series. Here, *x* is a vector containing the values for the x-axes and *y* is a matrix containing the data as a function of *x*. Note, that ```matplot(x, y, ...)``` plots each *column* of *y* as separate data series.

```{r eval=FALSE}
x <- 1:10 # creates a vector of length 10, this will become the x-axes
y.1 <- x^2 # another vectors of length 10
y.2 <- x^2 - 2*x # and a third vector of length 10

matplot(x, cbind(y.1, y.2), col =c(2, 4), pch = c(0, 2), xlab = "x", ylab = "y.1 = x^2 and y.2 = x^2-2x")
legend(1, 100, legend = c("y.1","y.2"), col =c(2, 4), pch = c(0, 2)) # adds a plot legend
```

In the following example, the variable *x* is given by the first column of a matrix while the actual data are taken from the second and third column of that same matrix.

```{r eval=FALSE}
x <- 1:10 # creates a vector of length 10, this will become the x-axes
y.1 <- x^2 # two more vectors of length 10
y.2 <- x^2 - 2*x

data_matrix <- cbind(x, y.1, y.2) # combine the above three vectors into a matrix
data_matrix

matplot(data_matrix[ , 1], data_matrix[ , 2-3], pch = c(16, 1), xlab = "x", ylab = "y.1 = x^2 and y.2 = x^2-2*x")
legend(1, 100, legend = c("y.1","y.2"), pch = c(16, 1), col = c(1, 2)) 
```

### Histograms

The function ```hist(x, ...)``` is used to produce simple histograms, where *x* is a vector of values for which the histogram is desired. Histograms show the abundance or frequency of data falling in certain discrete classes.

```{r eval = FALSE}
# generate data, drawn from Poisson distribution with mean 3
d1 <- rpois(100, lambda = 3)
# basic histogram
hist(d1) # without further options R decides how to bin the data
# specify desired number of bins
hist(d1, breaks = 4) # four break points results in five bins
# specify bin edges
hist(d1, breaks = c(0, 1, 3, 5, 7, 11, 21)) # manually defining bins
# use frequencies (default if bins have equal size)
hist(d1, freq = TRUE)
# use density (= relative frequencies)
hist(d1, freq = FALSE)
# get the histogram, but without plotting it
z <- hist(d1, plot = FALSE)
# access elements of the histogram
z$counts # counts per bin
z$mids # midpoints of the bins
```

### Bar Plots

A bar plot is used to represent data for discrete groups. Here are some examples using the function ```barplot(height, ...)```:

```{r eval = FALSE}
data(rivers) # load data on length of North American rivers
barplot(rivers, xlab = "rivers", ylab = "length")
barplot(rivers, xlab = "rivers", ylab = "length", horiz = TRUE) # horizontal
data(iris) # load iris data set
structure(iris) # inspect the iris data set
# color by group level
barplot(height = iris$Petal.Width, beside = TRUE, col = iris$Species, xlab = "species", ylab = "petal width")
```

### Box Plots

Box plots are used to show the range of a distribution, and the location of the bulk of its mass. Use the function ```boxplot(x, ...)``` where *x* specifies the data from which the boxplot is to be produced.

```{r eval = FALSE}
data(iris)
# set specific colors
boxplot(iris$Petal.Width ~ iris$Species, col = c("red", "green", "blue"))
```

