---
pagetitle: 'Learning Coding in R'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Coding in R {.tabset}

## Basic Plotting

In R, you can choose different ways to plot your data and results. By default, R ships with built-in graphical functions. Other functions, or entirely different paradigms, can be enabled using different packages. Here we give a brief overview of the standard plotting functions that are useful for quick data exploration. These are also referred to as R “base graphics” as no extra package is required.

R base graphics are drawn interactively: you can start with a plot, and then overlay other elements on the existing plot.

### Scatter Plots

The function ```plot(x, y, ...)``` is the most basic way to plot points defined by their x- and y-coordinates. Here, ```x``` and ```y``` are vectors of equal length containing the x- and y-coordinates of the point to be plotted. For example,

```{r eval=FALSE}
# generate data
x <- 1:30 # creates a vector of 30 values that we use as x-coordinates
y <- rnorm(30, mean = x) # creates a vector of 30 values that we use as y-coordinates; each values is drawn from normal distribution with with an increasing mean value
plot(y) # if no x-vector is given the defaul x = (1, 2, 3, ...) is used
plot(x, y) # this plot produces an identical result
# there are many different option one can add to the plot command
plot(x, y, type = "l") # using lines instead of points
plot(x, y, type = "b") # using both lines and points
plot(x, y, type = "b", pch = 4) # change the type of point
plot(x, y, type = "b", pch = 4, col = "blue") # change color
abline(a = 0, b = 1) # add a line to the same plot by specifying its intercept a and slope b,
abline(h = 20) # adding a horizontal line by setting a y-value h
abline(v = 10) # adding a vertical line by setting an x-value v
plot(x, y, col = "orange", xlab = "my x-label", ylab = "a y-label")
plot(x, y2, xlim = c(1,35), ylim = c(1,40)) # set plot ranges
```

### Function Plots

We can plot the graphs of functions using the command ```curve(expr, from, to, ...)```, where ```expr``` is a mathematical expression containing an argument ```x``` or function name, and ```from``` and ```to``` defines the range of ```x``` for which the function is plotted. This is a useful tool for visually exploring the shape of a function.

```{r eval=FALSE}
curve(exp, from = -3, to = 3) # plot of the exponential function in the interval (-3, 3)
abline(v = 0) # adding a vertical line at x = 0
logistic <- function(n, r, k){# definition of the user-defined function giving the population growth rate of a logistically growing population as a function the current population size n, the intrinsic growth rate r and the carrying capacity k
  n*r*(1-n/k)
}
curve(logistic(x , r = 1, k = 2), from = 0, to =2.5) # plot of the function logistic as a function of the population size n = x
abline(h = 0) # adding a horizontal line at y = 0
```

### Histograms

The function ```hist(x, ...)``` is used to produce simple histograms, where ```x``` is a vector of values for which the histogram is desired. Histograms show the abundance or frequency of data falling in certain discrete classes.

```{r eval = FALSE}
# generate data, drawn from Poisson distribution with mean 3
d1 <- rpois(100, lambda = 3)
# basic histogram
hist(d1) # without further options R decides how bin the data
# specify desired number of bins
hist(d1, breaks = 4) # four break points results in five bins
# specify bin edges
hist(d1, breaks = c(0, 1, 3, 5, 7, 11, 21)) # manually defining bins
# use frequencies (default if bins have equal size)
hist(d1, freq = TRUE)
# use density
hist(d1, freq = FALSE)
# get the histogram, but without plotting it
z <- hist(d1, plot = FALSE)
# access elements of the histogram
z$counts # counts per bin
z$mids # midpoints of the bins
```

### Bar Plots

A bar plot is used to represent data for discrete groups. Here are some examples using the function ```barplot(height, ...)```:

```{r eval = FALSE}
data(rivers) # load data on length of North American rivers
barplot(rivers, xlab = "rivers", ylab = "length")
barplot(order(rivers), xlab = "rivers", ylab = "length")
barplot(rivers, xlab = "rivers", ylab = "length", horiz = TRUE) # horizontal
data(iris) # load iris data set
structure(iris) # inspect the iris data set
# color by group level
barplot(height = iris$Petal.Width, beside = TRUE, col = iris$Species, xlab = "species")
```

### Box Plots

Box plots are used to show the range of a distribution, and the location of the bulk of its mass. Use the function ```boxplot(x, ...)``` where ```x``` specifies the data from which the boxplot is to be produced.

```{r eval = FALSE}
data(iris)
# set specific colors
boxplot(iris$Petal.Width ~ iris$Species, col = c("red", "green", "blue"))
```

