---
pagetitle: 'Learning Coding in R'
output: 
  html_document:
    theme: cerulean
    toc: no
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Solving differential equations in *R* {.tabset}

## Numerical integration and the *R*-package *deSolve*

Almost all ecological models formulated as *ordinary differential equations* (ODEs) cannot be solved explicitly. That is to say, no explicit formula can be derived that gives population size as a function of time (logistic growth is a notable exception, see below). However, ODE models can in general very easily be analyzed in *R* with a method called [*numerical integration*](https://en.wikipedia.org/wiki/Numerical_methods_for_ordinary_differential_equations). What does that mean?

The easiest method is called the *Euler method* and works as follows. Let us consider the ordinary differential equation

$$
\frac{\mathrm{d}x}{\mathrm{d}t} = f(x(t)),
$$

where $x$ could stand for population size and $f$ is a function describing the rate of change. From any starting population size $x_0$, we can find an approximation of the value of $x$ a short time $\Delta t$ later through the finite time difference equation

$$
\frac{\mathrm{d}x}{\mathrm{d}t} \approx \frac{x(t + \Delta t) − x(t)}{\Delta t},
$$
which can be rearranged as
$$
x(t + \Delta t) \approx x(t) + \Delta t * \frac{\mathrm{d}x}{\mathrm{d}t},
$$
or, using the the right-hand side of the differential equation, as
$$
x(t + \Delta t) \approx x(t) + \Delta t * f(x(t)).
$$
In words, the value of the dynamic variable $x$ a short time $\Delta t$ later can be approximated by the current value $x(t)$ plus the slope of the curve, given by $f(x(t))$, multiplied by the length of the time interval $\Delta t$. Thus, we locally approximate the function $x(t)$ with its tangent line. 

This formula is usually applied in the following way. We choose a step size $\Delta t$, and then construct the sequence $t_0,\, t_1 = t_0 + \Delta t,\, t_2 = t_0 + 2\Delta t,\, \ldots$. We denote by $x_n$ a numerical estimate of the exact solution $x(t_n)$, which are computed using the recursive scheme
$$
x_{n+1} = x_n + \Delta t * f(x_n).
$$
The smaller the length of the time interval $\Delta t$, the better the approximation, and $\Delta t$ will indeed be an important parameter when solving ODEs in the next section. The Euler method is the easiest method to numerically solve an ordinary differential equation but far more accurate methods exist and are implemented in the package *deSolve*.

### The *R*-package *deSolve*

The *R*-package *deSolve* allows us to numerically solve ecological ODE models. A detailed documentation for this package and tutorials can be found [here](https://desolve.r-forge.r-project.org/). Start by installing the package via the *Install*-tab in the bottom-right of your *RStudio* window and load it using

```{r}
library(deSolve)
```

As a first example, let us use the model for logistic growth in the $r/\alpha$ formulation:

$$
\frac{\mathrm{d}N}{\mathrm{d}t}=N \left(r  - \alpha N \right)
$$
This model has the stable equilibrium $\hat N = r/\alpha$, often referred to as carrying capacity. In order to analyze an ODE, we need to do four things:

1) Specify the initial values of the dynamics variables.
 
```{r}
N0 <- c(N = 1)
```

2) Specify the parameters of the model

```{r}
parameters <- c(r = 1, a = 0.01)
```

3) Specify the vector of time-steps at the which the ODE will be evaluated.

```{r}
time.steps <- seq(0, 10, 2) # this creates the vector (0, 2, 4, 6, 8, 10). Thus, here Delta t equals 1.
```

4) Specify a function containing the differential equation model.

```{r}
log.growth <- function(time.steps, N0, parameters) {
  with(as.list(c(N0, parameters)), {
    dN.dt <- N * r*(1 - a * N)
    return(list(dN.dt))
  })
}
```

Note that we here stored the parameters and initial conditions in a *named vector*, which allows us to call their entries by their name. The *R*-statement with ```with(as.list(c(y, parms)), { }``` ensures that the state variables and parameters can be addressed by their names.

Now we are ready to apply the *deSolve*-function ```ode(y, times, func, parms, ...)```. This function has four obligatory arguments, the initial values *y*, the vector of time steps *times*, the name of the function that is going to be solved *func*, and the vector of parameters *parms*. The output is a data frame, with the first column being the time steps, and the remaining being the state variables.

```{r}
time.series <- ode(y = N0, times = time.steps, func = log.growth, parms = parameters)
time.series # let us have a look at the resulting data
```

```{r}
plot(time.steps, time.series[, 2], type = "l", lty = 1:2, col=1, ylab = "Population Density N", xlab="Time")
```

This graph shows the solution of the ODE for logistic growth. We can make two observations. First, it seems that the dynamics might not have quite reached its equilibrium (given by $\hat N = r/a = 1/0.01 = 100$). We therefore should run the model a bit longer, in other words, increase the length of the time as specified in *time.steps*. Second, the graph of the time series does not look smooth but rather a bit choppy. This is due to the fact that we calculate the solution at only 6 different points, as determined by *time.steps <- seq(1, 10, 2)*. Let us quadruple the number of time points at which the solution is calculated.

```{r}
time.steps.2 <- seq(0, 15, 0.5) # this creates the vector (0, 0.5, 1.5, 2, 2.5, ..., 15). Thus, here Delta t equals 0.5.
time.series.2 <- ode(N0, time.steps.2, log.growth, parameters)

plot(time.steps.2, time.series.2[, 2], type = "l", lty = 1, col=1, ylab = "Population Density N", xlab="Time") 
lines(time.steps, time.series[, 2], type = "l", lty = 2, col=2)
legend("topleft", legend = c("high resolution","low resolution"),  col =c(1, 2), lty = c(1, 2))
```

And indeed, the new curve looks very smooth. For the logistic model it is in fact possible to compute the solution *analytically*, 

$$
N(t) = \frac{N_0 \exp(r t)}{1 + \alpha r^{-1} N_0 (\exp(r t) - 1)},
$$

which we can define as an *R*-function and plot using the function *curve*:

```{r eval=FALSE}
solution_logistic <- function(t) { (N0 * exp(r * t)) /
    (1 + (a * N0 / r) * (exp(r * t) - 1))
}

r <- 1
a <- 0.01
N0 <- 1

curve(solution_logistic(x), from = 0, to = 15, xlab = "time", ylab = "population size N")
```

This graph is virtually indistinguishable from the one based on numerical integration.

## Solving systems of coupled ODEs

Many ecological models describe the rate of change of more than one variable, such as a predator and its prey, or two competing species. The package *deSolve* can also solve such *systems of coupled ordinary differential equations*. As case in point, let us analyze the continuous time Lotka-Volterra competition model, which can be written as

$$
\frac{\mathrm{d}N_1}{\mathrm{d}t} = N_1 \left(r_1 - \alpha_{11} N_1 - \alpha_{21} N_2 \right)\\
\frac{\mathrm{d}N_2}{\mathrm{d}t} = N_2 \left(r_2 − \alpha_{22} N_2 - \alpha_{21} N_1 \right),
$$
where $\alpha_{ij}$ describes the sensitivity of competition of species $i$ to the presence of species $i$.

To solve this system of equations numerically, we again start by specifying (i) the initial population sizes, (ii) the parameter values and (iii) the sequence of time steps for which to calculate population sizes.

```{r}
N <- c(N1 = 5, N2 = 1) # vector of initial population sizes
r <- c(r1 = 1, r2 = 1) # vector of r-values
a <- c(a11 = 0.01, a12 = 0.008, a22 = 0.009, a21 = 0.007) # vector of sensitivities
parameters <- list(r, a) # all parameters collected in one list
time.steps <- seq(1, 50, 0.5) # sequence of times steps
```

Next, we specify the function with the two differential equations.

```{r}
LV.comp <- function(time.steps, N, parameters) {
  with(as.list(c(N, parameters)), {
  dN1.dt <- N[1] * (r[1] - a[1] * N[1] - a[2] * N[2]) # here [i] indicates the ith position in the corresponding vector
  dN2.dt <- N[2] * (r[2] - a[3] * N[2] - a[4] * N[1])
  return(list(c(dN1.dt, dN2.dt)))
  })
}  
```

Note that *LV.comp* assumes that *N*, *r*, *a* and *K* are vectors. For instance, the function extracts the first entry of *r* for the first species *(r[1])*. The vector *N* of population sizes contains one value for each population. To solve the population dynamics, we again use the function *ode* just as in the previous example.

```{r}
time.series <- ode(y = N, times = time.steps, func = LV.comp, parms = parameters)
time.series[1:5, ] # shows the two densities for the first five time steps
```

```{r}
matplot(time.steps, (time.series[, 2:3]), type = "l", lty = 1:2, col=1, ylab = "Population Density N", xlab="Time")
legend("bottomright", c("species 1", "species 2"), lty = 1:2)
```

For the above parameter values, the two competing species can coexist with each other. It seems that the sequence of time steps is both long enough and sufficiently fine grained for the solution to look smooth and reach its equilibrium.

Instead plotting these simulations as a time series, we can also plot them as a phase-plane diagram.

```{r}
plot(time.series[, 2], time.series[, 3], xlim = c(0, 80), ylim = c(0, 100), type = "l",  lty = 1, xlab = "species 1", ylab = "species 2")
points(N[1], N[2], cex = 1.5, pch = 19, col = "red") # adds a dot to show the starting densities
```

<!---
STUFF TO ADD IN THE FUTURE
The ```ode()``` function uses a superb ODE solver, ```lsoda```, which is a very powerful, well tested tool, superior to many other such solvers. One of the other solvers in the ```deSolve```-package, ```lsodar```, will also return roots (or equilibria), for a system of ODEs, if they exist. Here we find the roots (i.e. the solutions, or equilibria) for a two species competition model.

To use ```lsodar``` to find equilibria, we need to specify a root finding function whose inputs are are the sme of the ODE function, and which returns a scalar (a single number) that determines whether the rate of change (dy/dx) is sufficiently close to zero that we can say that the system has stopped changed, that is, has reached a steady state or equilibrium. Here we sum the absolute rates of change of each species, and then subtract 10−10; if that difference is zero, we decide that, for all pratcial purposes, the system has stopped changing.

#```{r}
#rootfun <- function(t, y, p) {
#  dstate <- unlist(LV_time_series(t, y, p))
#  return(sum(abs(dstate)) - 1e-10)
#}
#```

Note that ```unlist``` changes the ```list``` returned by EV into a simple vector, which can then be summed.

Next we specify parameters, and time. Here all we want is the root, so we specify that we want the value of y after a really long time (t = 1010). The ```lsodar``` function will stop sooner than that, and return the equilibrium it finds, and the time step at which it occurred.

#```{r}
#p <- c(b = 0.5, a = 0.02, e = 0.1, s = 0.2)
#t <- c(0, 1e+10)
#```

Now we run the function.

#```{r}
#out <- ode(y = c(45, 200), t, LV_time_series, parms = p, rootfun = #rootfun, method = "lsodar")
#out[, ]
#```

Here we see that the steady state population sizes are V = 100 and E = 12.5, and that given our starting point, this steady state was achieved at t = 500.8. Other information is available; see ?lsodar after loading the deSolve package.
--->
