---
pagetitle: 'Learning Coding in R'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, eval=FALSE, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# *R* as an Oversized Calculator: the Console {.tabset}

## Getting started

*R* is a versatile programming language for statistical analysis, data analysis and numerical calculations. The main hurdle new users face when learning *R* is that it is based on a command-line interface: to make things happen, you write text commands in a *console*, and then the program executes them. This might seem unusual if you are used to software based on a graphical user interface, where you tend to work by clicking on windows and buttons. However, the command line is what makes it easy to automate your analysis - all you have to do is collect all the commands in a text file, and then run them in *R*.

### Installing *R* and *RStudio*

For this introduction, we are going to use *RStudio*, a graphical interface that simplifies the use of *R* by giving you immediate access to the code, the console, and the graphics.

To install *R* and *RStudio*, you can follow [these instructions](https://rstudio-education.github.io/hopr/starting.html).

### Finding Help

Each command in *R* comes with a manual page. To access it, type ```?[NAME OFCOMMAND]``` in the console (e.g., ```?lm```). Alternatively, search the command under the help-tab in the bottom right window of RStudio.

But really, if you need help with *R* the internet is your friend. Virtually any problem you might encounter has been described and solved somewhere on the internet. It's all about to phrase your question in the right manner in a Google search. Most often you will end up on [https://stackexchange.com/](https://stackexchange.com/), [https://stackoverflow.com](https://stackoverflow.com/questions) or [https://www.r-bloggers.com/](https://www.r-bloggers.com/).

## The Console

We will first explore simple operations, assignments, and data types to get familiar with the *R* console. A "greater than sign" (>) at the **beginning of the line in the console** means that *R* is ready to accept your input. You can navigate the history of previously typed commands by using the arrows on your keyboard. A comment starts with the hash (#), and does not require a closing symbol.

We can use *R* as an oversized calculator:

```{r comment="", prompt=TRUE, eval=FALSE}
1.7 * 2 # multiplication
12 / 5 # division
2.1 ^ 5 # power
log(10) # natural logarithm
log10(10) # logarithm to base 10
sqrt(9) # square root
```

The following table provides a list of arithmetic and logical operators in R:

Operators | Description
----------|------------
+ | Addition
- | Subtraction
* | Multiplication
/ | Division
^ or ** |  Exponentiation
== | Equal to
!= | Differs from
> | Greater than
< | Less than
>= | Greater than or equal to
<= | Less than or equal to
& | Logical and
&#124; | Logical or
! | Logical not

Let’s have a closer look at the logical operators. They create a logical value by comparing two elements:
 
```{r comment="", prompt=TRUE, eval=FALSE}
5 > 3
5 == (10 / 2)
6>2^4
6 >= (2 * 3)
(5 > 3) & (7 < 5) # logical AND
(5 > 3) | (7 < 5) # logical OR
```

*R* is designed for statistical analysis, and therefore it provides many built-in mathematical functions. Some are listed in the next table.

Function | Description
---------|------------
abs(x) | Absolute value
sqrt(x) | Square root
cos(x), sin(x), tan(x) | Trigonometric functions
log(x) | Natural logarithm
log10(x) | Base 10 logarithm
exp(x) | eponential function $e^x$

We can use the operator ```%in%``` to search for matches. It returns a logical value (i.e., ```TRUE``` or ```FALSE```) indicating whether or not a match exists.

```{r comment="", prompt=TRUE, eval=FALSE}
3 %in% 1:5 # 1:5 generates a sequence
# c() combines values into a vector or list
# all arguments are converted to the same type (here it is a character string)
c(2, "good to be", TRUE)
2 %in% c(2, "good to be", TRUE)
# test for multiple elements at once
1:8 %in% c(1, 3, 5)
```

## Assignment and Data Types

Everything in *R* is an *object*, where an object is a name that you can use to call up stored data.  When programming in R, you assign data to objects using the assignment command is ```<-``` (less-than sign, followed by a minus sign). Each new object will appear in the *environment*-pane in the top-right of *RStudio*.

```{r comment="", prompt=TRUE, eval=FALSE}
x <- 5 # assign the value 5 to x
x * 2 # use variable x to perform operations
x <- 7 # assign new value to variable, this overwrites the previous assignment
x*2
```

Note that the previous value is overwritten whenever a new value is assigned to the variable. To list all the variables you have created, type ```ls()```. To remove a variable *x* from the current *R* session, type ```rm(x)```.

*R* can handle different types or modes of data, the most important ones are *numeric* (numbers), *character* (strings of character, e.g., a word), and *logical* (TRUE and FALSE). In the following, we assign a new value to the variable *x*, each time altering its *type*. To determine the type of variable ```x```, use the command ```class```. You can also test whether a variable is of a certain type by using the functions ```is.numeric```, ```is.character```, etc.

```{r comment="", prompt=TRUE, eval=FALSE}
# numeric (real numbers)
x <- 5
class(x)
X # note that R is cass sensitive, that is, x is different from X
x <- pi # a numeric that R knows
class(x)
# logical (TRUE or FALSE)
x <- (5 > 7)
class(x)
# character (strings)
x <- "hello"
class(x)
is.numeric(x)
```

If you need to change the type of a variable (an operation called *casting*), use the functions ```as.character``` or ```as.numeric```, etc. In *RStudio*, if you type the first few letters of the name of a function or a variable, you can see all the possible ways to complete the name; try typing ```as.``` to see all the available functions for type conversion. Some examples:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- "05.2" # assign a character string
class(x)
(x <- as.numeric(x)) # cast character string to numeric, round brackets around assignment produce output
as.logical(x) # all x not equal 0 are TRUE
as.logical(0) # only 0 is FALSE
```

## Objects

Everying in *R* is an *object*. Objects come in different types, the most important ones are *vectors*, *matrices*, *lists*, and *functions*.

### Vectors

The most basic data structure in *R* is the vector, which is an ordered collection of values of the same type. As such, vectors are *1-dimensional* objects. Vectors can be created by *concatenating* different values with the command ```c```:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- c(2, 3, 5, 2, 3, 13, 17, 3) # a numeric vector
y <- c("a", "b", "c") # a vector of characters
z <- c(1, "a") # entries in a vector have to be of the same type
z # R coerces 1 into a charater
class(z)
```

You can access the elements of a vector by their index. You can access the elements of a vector by specifying their positions:

```{r comment="", prompt=TRUE, eval=FALSE}
x[3]
x[8]
x[9] # what if the element does not exist?
```

Alternatively, you can access the elements of a vector by specifying using the logical ```TRUE```:
```{r comment="", prompt=TRUE, eval=FALSE}
x[c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)]
x[c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)]
```

You can extract several elements at once (i.e., create another vector) using the colon (```:```) command, or by concatenating the indices:

```{r comment="", prompt=TRUE, eval=FALSE}
x[1:3]
x[4:7]
x[c(1,3,5)]
```

Sometimes one wants to know at which position in a vector a certain values occurs. We can find out with the ```which``` function.

```{r comment="", prompt=TRUE, eval=FALSE}
which(x == 3) # the value 3 occurs at postion 2, 4 and 8
which(x > 5) # values larger than 5 occur at position 6 and 7
length(which(x == 3)) # the value 3 occurs exactly three times within the vector x
```

Given that *R* was born for statistics, there are many statistical functions you can perform on vectors:

```{r comment="", prompt=TRUE, eval=FALSE}
length(x)
min(x)
max(x)
sum(x) # sum all elements
prod(x) # multiply all elements
median(x) # median value
mean(x) # arithmetic mean
var(x) # unbiased sample variance
mean(x ^ 2) - mean(x) ^ 2 # population variance
summary(x) # print a summary
```

You can generate vectors of sequential numbers using the colon
command:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- 1:10
x
```

For more complex sequences, use ```seq```:

```{r comment="", prompt=TRUE, eval=FALSE}
seq(from = 1, to = 5, by = 0.5)
```

To repeat a value or a sequence several times, use ```rep```:

```{r comment="", prompt=TRUE, eval=FALSE}
rep("abc", 3)
rep(c(1,2,3), 3)
```

#### Exercises

(a) Create a vector containing all the even numbers between 2 and 100 (inclusive) and store it in the variable ```z```.
(b) What is the sum of all the elements of ```z```?
(c) Is the sum of the elements of ```z``` equal to 51 × 50?
(d) What is the product of entries at position 5, 10, and 15 of ```z```?
(e) Does ```seq(2, 100, by = 2)``` produce the same vector as ```(1:50) * 2```?
(f) What happens if you type ```z^2```?

<!--(b) Extract all the elements of ```z``` that are divisible by 12. How many elements match this criterion?
(f) Create a vector ```y``` that contains all numbers between 0 and 30 that are
divisible by 3. Find the five elements of ```y``` that are also elements of ```z```.-->

### Matrices

A ```matrix``` is a two-dimensional object where data are stored in a rectangular array with rows and columns. In the case of numeric values, you can perform common matrix operations (e.g., product, inverse, decomposition):

```{r comment="", prompt=TRUE, eval=FALSE}
# create matrix by specifying values, number of rows, number of columns
A <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2)
A # by default, the matrix is filled column by column
(A <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2, byrow = TRUE)) # fill matrix by row
A %*% A # matrix product
solve(A) # matrix inverse
A %*% solve(A)
diag(A) # vector containing the diagonal elements
B <- matrix(1, 3, 2) # 3 x 2 matrix filled with 1s
B
t(B) # transpose
B %*% t(B)
# by defaul, matrices are filled by column
(C <- matrix(1:9, 3, 3))
(D <- matrix(c("a", "b", "c", "d"), 2, 2))
A %*% D # matrix operations are not defined for matrices with numeric entries
```

Determine the dimensions of a matrix (the numbers of rows and
columns) using the following commands:

```{r comment="", prompt=TRUE, eval=FALSE}
dim(B)
nrow(B)
ncol(B)
```

Use indices to access a particular row/column of a matrix:

```{r comment="", prompt=TRUE, eval=FALSE}
Z
Z[1, ] # first row
Z[, 2] # second column
# select submatrix with coefficients of second
# and third columns of the first two rows
Z[1:2, 2:3]
# concatenate positions to index nonadjacent rows/columns
Z[c(1,3), c(1,3)]
```

Several functions exist to compute sums and averages of matrices:

```{r comment="", prompt=TRUE, eval=FALSE}
rowSums(Z) # sum of the values in each row
rowMeans(Z) # arithmetic mean for each row
colSums(Z) # sum of the values in each column
colMeans(Z) # arithmetic mean for each column
sum(Z) # sum of all matrix entries
mean(Z) # arithmetic mean over all matric entries
```

### Lists

Vectors are a good choice when each element is of the same type (e.g., numbers, strings). Lists are used when we want to store elements of different types, or more complex objects (e.g., vectors, matrices, even lists of lists). Each element of the list can be referenced either by its index, or by a label:

```{r comment="", prompt=TRUE, eval=FALSE}
# create list containing two vectors that are labelled Names and Values
mylist <- list(Names = c("a", "b", "c", "d"), Values = c(1, 2, 3))
mylist # prints the content of mylist
mylist[[1]] # access first entry within list by index
# note that in order to access entries in a list we have to use [[]]
mylist[[2]] # access second entry by index
mylist$Names # access first entry by label
mylist[["Names"]] # an alternative way to access by label
mylist[["Values"]][3] # third element of the second entry
```

### Functions

Functions are *R* objects designed to perform operations on other objects. Each time you write a command at the *R* prompt and hit the Enter-key, one or several functions are executed. An *R* function is typically called by its name followed by parentheses. Inside the parentheses, the user must pass the arguments required by the function. We have already described several built-in functions previously, such as ```c()```, ```log()```, and ```matrix()```. What makes *R* a powerful programming language is that you can define your own functions, something we will do later.

### Data Frames

Data frames are special matrices that contain additional information about its content such as row and column names. In short, data frames are *R*s equivalent to a an EXCEL preadsheet. The columns (typically representing different measurements) can be of different types (e.g., one column could be the date of measurement, another the weight of the individual, or the volume of the cell, or the treatment of the sample), while the rows typically represent different samples.

When you import a spreadsheet file into R, it is automatically stored as a data frame. The difference between a matrix and a data frame is that in a matrix all the values are of the same type (e.g., all numeric), while in a data frame each column can be of a different type.

Because typing a data frame by hand would be tedious, let’s use a data set that is already available in R:

```{r comment="", prompt=TRUE, eval=FALSE}
data(trees) # load data set "trees"
str(trees) # structure of data frame ’data.frame’
ncol(trees) # number of columns
nrow(trees) # number of rows
head(trees) # print the first few rows Girth Height Volume
trees$Girth # select column by name
# select column by name; return first 5 elements
trees$Height[1:5]
trees[1:3, ] # select rows 1 through 3
# select rows 1 through 3; return column Volume
trees[1:3, ]$Volume
trees <- rbind(trees, c(13.25, 76, 30.17)) # add a row
trees_double <- cbind(trees, trees) # combine columns
head(trees_double)
# change column names
colnames(trees) <- c("girth", "height", "volume")
head(trees)
```

#### Exercises

(a) What is the average tree height?
(b) What is the average girth of those trees that are more than 75 ft tall?
(c) What is the maximum height of trees with a volume between 15
and 35 ft^3?

