---
pagetitle: 'Learning Coding in R'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Coding in R {.tabset}

## Getting started

R is a versatatile programming language for statistical analysis, data analysis and numerical calculationns. The main hurdle new users face when learning R is that it is based on a command-line interface: to make things happen, you write text commands in a “console,” and then the program executes them. This might seem unusual if you are used to software based on a graphical user interface, where you tend to work by clicking on windows and buttons. However, the command line is what makes it easy to automate your analysis - all you have to do is collect all the commands in a text file, and then run them in R.

### Installing R and RStudio

For this introduction, we are going to use RStudio, a graphical interface that simplifies the use of R by giving you immediate access to the code, the console, and the graphics.

To install R and RStudio, you can follow [these instructions](https://rstudio-education.github.io/hopr/starting.html).

### Finding Help

Each command in R comes with a manual page. To access it, type ```?[NAME OFCOMMAND]``` in the console (e.g., ```?lm```). Alternatively, search the command under the help-tab tin the bottom right window of RStudio.

But really, if you need help with R the internet is your friend. Virtually any problem you might encounter has been described and solved somewhere on the internet. It's all about to phrase your question in the right manner in a Google search. Most often you will end up on [https://stackexchange.com/](https://stackexchange.com/), [https://stackoverflow.com](https://stackoverflow.com/questions) or [https://www.r-bloggers.com/](https://www.r-bloggers.com/).

## The Console

We will first explore simple operations, assignments, and data types to get familiar with the R console. A "greater than sign" (>) at the **beginning of the line in the console** means that R is ready to accept your input. You can navigate the history of previously typed commands by using the arrows on your keyboard. A comment starts with the hash (#), and does not require a closing symbol.

We can use R as an oversized calculator:

```{r comment="", prompt=TRUE, eval=FALSE}
1.7 * 2
12 / 5
2.1 ^ 5
log(10)
log10(10)
sqrt(9)
```

The following table provides a list of arithmetic and logical operators in R:

Operators | Description
----------|------------
+ | Addition
- | Subtraction
* | Multiplication
/ | Division
^ or ** |  Exponentiation
== | Equal to
!= | Differs from
> | Greater than
< | Less than
>= | Greater than or equal to
<= | Less than or equal to
& | Logical and
&#124; | Logical or
! | Logical not

R is designed for statistical analysis, and therefore it provides many built-in mathematical functions. Some are listed in the next table.

Function | Description
---------|------------
abs(x) | Absolute value
sqrt(x) | Square root
cos(x), sin(x), tan(x) | Trigonometric functions
log(x) | Natural logarithm
log10(x) | Base 10 logarithm
exp(x) | eponential function $e^x$

Let’s have a closer look at the logical operators. They create a logical value by comparing two elements:
 
```{r comment="", prompt=TRUE, eval=FALSE}
5 > 3
5 == (10 / 2)
6>2^4
6 >= (2 * 3)
(5 > 3) & (7 < 5) # logical AND
(5 > 3) | (7 < 5) # logical OR
```

We can use the operator ```%in%``` to search for matches. It returns a logical value (i.e., ```TRUE``` or ```FALSE```) indicating whether or not a match exists.

```{r comment="", prompt=TRUE, eval=FALSE}
3 %in% 1:5 # 1:5 generates a sequence
# c() combines values into a vector or list
# all arguments are converted to the same type
c(2, "good to be", TRUE)
2 %in% c(2, "good to be", TRUE)
# test for multiple elements at once
1:8 %in% c(1, 3, 5)
```

## Assignment and Data Types

When programming in R, you assign values to variables: a variable is a "box" that can contain a value or object. The assignment command in R is ```<-``` (less-than sign, followed by a minus sign).

```{r comment="", prompt=TRUE, eval=FALSE}
x <- 5 # assign the value 5 to x
x * 2 # use variable x to perform operations
x <- 7 # assign new value to variable >x*2
```

Note that the previous value is overwritten whenever a new value is assigned to the variable. To list all the variables you have created, type ```ls()```. To remove a variable ```x``` from the current R session, type ```rm(x)```.

R can handle different types of data. In the following, we assign a new value to the variable ```x```, each time altering its *data type*. To determine the type of variable ```x```, use the command ```class```. You can also test whether a variable is of a certain type by using the functions ```is.numeric```, ```is.character```, etc.

```{r comment="", prompt=TRUE, eval=FALSE}
# integer (natural numbers)
x <- as.integer(5)
class(x)
# numeric (real numbers)
x <- pi
class(x)
# complex (complex number)
x <- 1 + 3i
class(x)
# logical (TRUE or FALSE)
x <- (5 > 7)
class(x)
# character (strings)
x <- "hello"
class(x)
is.numeric(x)
```

If you need to change the type of a variable (an operation called *casting*), use the functions ```as.character```, ```as.integer```, ```as.numeric```, etc. In RStudio, if you type the first few letters of the name of a function or a variable, and hit Tab, you can see all the possible ways to complete the name; try typing ```as.``` and hit Tab to see all the available functions for type conversion. Some examples:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- 5
as.character(x)
as.logical(x) # only 0 is FALSE
y <- "07.123" # assign a character string
x < y # beware: comparing different types
x < as.numeric(y) # cast string to numeric
```

## Data Structures

R ships with several data structures, which can be used to organize your data. Knowing the characteristics and specific operations of each data structure allows you to write better and more compact code.

### Vectors

The most basic data structure in R is the vector, which is an ordered collection of values of the same type. Vectors can be created by concatenating different values with the command ```c```:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- c(2, 3, 5, 2, 3, 13, 17, 3)
```

You can access the elements of a vector by their index. You can access the elements of a vector by specifying their positions:

```{r comment="", prompt=TRUE, eval=FALSE}
x[3]
x[8]
x[9] # what if the element does not exist?
```

You can extract several elements at once (i.e., create another vector) using the colon (```:```) command, or by concatenating the indices:

```{r comment="", prompt=TRUE, eval=FALSE}
x[1:3]
x[4:7]
x[c(1,3,5)]
```

Sometimes one wants to know at which position in a vector a certain values occurs. We can find out with the ```which``` function.

```{r comment="", prompt=TRUE, eval=FALSE}
which(x == 3) # the value 3 occurs at postion 2, 4 and 8
which(x > 5) # values larger than 5 occur at position 6 and 7
length(which(x == 3)) # the value 3 occurs exactly three times within the vector x
```

Given that R was born for statistics, there are many statistical functions you can perform on vectors:

```{r comment="", prompt=TRUE, eval=FALSE}
length(x)
min(x)
max(x)
sum(x) # sum all elements
prod(x) # multiply all elements
median(x) # median value
mean(x) # arithmetic mean
var(x) # unbiased sample variance
mean(x ^ 2) - mean(x) ^ 2 # population variance
summary(x) # print a summary
```

You can generate vectors of sequential numbers using the colon
command:

```{r comment="", prompt=TRUE, eval=FALSE}
x <- 1:10
x
```

For more complex sequences, use ```seq```:

```{r comment="", prompt=TRUE, eval=FALSE}
seq(from = 1, to = 5, by = 0.5)
```

To repeat a value or a sequence several times, use ```rep```:

```{r comment="", prompt=TRUE, eval=FALSE}
rep("abc", 3)
rep(c(1,2,3), 3)
```

#### Exercises

(a) Create a vector containing all the even numbers between 2 and 100 (inclusive) and store it in the variable z.
(b) Extractalltheelementsofzthataredivisibleby12.Howmanyelements match this criterion?
(c) What is the sum of all the elements of z?
(d) Is the sum of the elements of z equal to 51 × 50?
(e) What is the product of elements 5, 10, and 15 of z?
(f) Create a vector y that contains all numbers between 0 and 30 that are
divisible by 3. Find the five elements of y that are also elements of z.
(g) Does seq(2, 100, by = 2) produce the same vector as (1:50) * 2?
(h) What happens if you type z ^ 2?

### Matrices

A matrix is a two-dimensional table of values. In the case of numeric values, you can perform common matrix operations (e.g., product, inverse, decomposition):

```{r comment="", prompt=TRUE, eval=FALSE}
# create matrix
# indicate values, number of rows, number of columns
A <- matrix(c(1, 2, 3, 4), 2, 2)
A
A %*% A # matrix product
solve(A) # matrix inverse
A %*% solve(A)
diag(A) # vector containing the diagonal elements
B <- matrix(1, 3, 2) # 3 x 2 matrix filled with 1s
B
t(B) # transpose
B %*% t(B)
# by defaul, matrices are filled by column
Z <- matrix(1:9, 3, 3)
Z
# to fill by rows specify matrix (1:9, 3, 3, byrow = TRUE)
```

Determine the dimensions of a matrix (the numbers of rows and
columns) using the following commands:

```{r comment="", prompt=TRUE, eval=FALSE}
dim(B)
nrow(B)
ncol(B)
```

Use indices to access a particular row/column of a matrix:

```{r comment="", prompt=TRUE, eval=FALSE}
Z
Z[1, ] # first row
Z[, 2] # second column
# select submatrix with coefficients of second
# and third columns of the first two rows
Z[1:2, 2:3]
# concatenate positions to index nonadjacent rows/columns
Z[c(1,3), c(1,3)]
```

Most functions consider all coefficients in the matrix:

```{r comment="", prompt=TRUE, eval=FALSE}
sum(Z)
mean(Z)
```

### Lists

Vectors are a good choice when each element is of the same type (e.g., numbers, strings). Lists are used when we want to store elements of dif- ferent types, or more complex objects (e.g., vectors, matrices, even lists of lists). Each element of the list can be referenced either by its index, or by a label:

```{r comment="", prompt=TRUE, eval=FALSE}
# create list containing two vectors that are labelled Names and Values
mylist <- list(Names = c("a", "b", "c", "d"), Values = c(1, 2, 3))
mylist # prints the content of mylist
mylist[[1]] # access first entry within list by index
# note that inorder to access entries in a list we have to use [[]]
mylist[[2]] # access second entry by index
mylist$Names # access first entry by label
mylist[["Names"]] # an alternative way to access by label
mylist[["Values"]][3] # third element of the second entry
```


### Data Frames

Data frames contain data organized like in a spreadsheet. The columns (typically representing different measurements) can be of different types (e.g., one column could be the date of measurement, another the weight of the individual, or the volume of the cell, or the treatment of the sample), while the rows typically represent different samples.

When you import a spreadsheet file into R, it is automatically stored as a data frame. The difference between a matrix and a data frame is that in a matrix all the values are of the same type (e.g., all numeric), while in a data frame each column can be of a different type.

Because typing a data frame by hand would be tedious, let’s use a data set that is already available in R:

```{r comment="", prompt=TRUE, eval=FALSE}
data(trees) # load data set "trees"
str(trees) # structure of data frame ’data.frame’
ncol(trees) # number of columns
nrow(trees) # number of rows
head(trees) # print the first few rows Girth Height Volume
trees$Girth # select column by name
# select column by name; return first 5 elements
trees$Height[1:5]
trees[1:3, ] # select rows 1 through 3
# select rows 1 through 3; return column Volume
trees[1:3, ]$Volume
trees <- rbind(trees, c(13.25, 76, 30.17)) # add a row
trees_double <- cbind(trees, trees) # combine columns
# change column names
colnames(trees) <- c("girth", "height", "volume")
```

#### Exercises

(a) What is the average height of the cherry trees?
(b) What is the average girth of those that are more than 75 ft tall?
(c) What is the maximum height of trees with a volume between 15
and 35 ft3?

